Bayes Computational Methods

### 2020-05-20 Paper.Rmd
Paper updates

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = FALSE)
```

@@

In Bayesian statistical inference, we are interested in finding the posterior distribution of a certain parameter, given a prior distribution and a likelihood for that parameter from a sample of data. Bayes Rule states that, given a parameter of interest $\theta$ and some known data $x$, we can find the conditional probability of $\theta$ given $y$ using $$ p(\theta | x) = \frac{p(x | \theta) p(\theta)}{p(x)},$$ where in the discrete case $p(x) = \sum_\theta p(\theta) p(x|\theta),$ and in the continuous case, $p(x) = \int p(\theta) p(y|\theta) d\theta$. More formally, we can derive the posterior distribution of a parameter $\theta$, $p(\theta | x)$, by $$p(\theta | x_1, x_2, \dots, x_n) = \frac{\ell(\theta | x_1, x_2, \dots, x_n) \pi(\theta)}{\int{\ell(\theta | x_1, x_2, \dots, x_n) \pi(\theta) d \theta}},$$ where $\ell(x | \theta)$ is proportional to the likelihood of $\theta$, conditioned on the data, and $\pi(\theta)$ is the prior distribution of $\theta$. For example, suppose that $X_i \sim \text{Bern}(\theta)$, and we are interested in inference on $\theta$. A reasonable prior to assign to $\theta$ would be a $\text{Beta}(\alpha, \beta)$, distribution, given that $\theta$ is constrained to be between 0 and 1, which is a property of the Beta distribution. Therefore, given that $\theta \sim \text{Beta}(\alpha, \beta)$ for some known values of $\alpha$ and $\beta$, then 
\begin{align*}
  p(\theta | x) &= \frac{\ell(\theta | x_1, x_2, \dots, x_n) \pi(\theta)}{\int{\ell(\theta | x_1, x_2, \dots, x_n) \pi(\theta) d \theta}} \\
  &= \frac{\theta ^ {\sum_{i=1}^{n} x_i} (1 - \theta) ^ {n - \sum_{i=1}^{n} x_i} \frac{\Gamma (\alpha + \beta)}{\Gamma(\alpha) + \Gamma(\beta)} \theta^{\alpha - 1} (1 - \theta)^{\beta - 1}}{\int \theta ^ {\sum_{i=1}^{n} x_i} (1 - \theta) ^ {n - \sum_{i=1}^{n} x_i} \frac{\Gamma (\alpha + \beta)}{\Gamma(\alpha) + \Gamma(\beta)} \theta^{\alpha - 1} (1 - \theta)^{\beta - 1} d \theta} \\
  &\propto \theta ^ {\sum (x_i) + \alpha - 1} (1 - \theta) ^ {n - \sum (x_i) + \beta - 1},
\end{align*}
which is the kernel of a $\text{Beta}\left(\sum (x_i) + \alpha, n - \sum (x_i) + \beta \right)$ distribution. Therefore, we can analytically calculate the posterior distribution of $\theta$ as $$p(\theta | x) \sim \text{Beta}\left(\sum (x_i) + \alpha, n - \sum (x_i) + \beta\right).$$

## Bayesian Computation

In more complicated scenarios, it is not easy to calculate the posterior distribution of a parameter by hand, and in many cases it is impossible. This is where computing can help us. If we could somehow computationally grab samples from the posterior distribution we are interested in, then we could obtain most of the statistics that we are interested in.

### History of Bayesian computation up to MCMC

@@

Markov Chain Monte Carlo (referred to as MCMC), is a common approach that allows for drawing samples of a parameter from a distribution that is approximately equivalent to the target posterior distribution of that parameter. It is an iterative algorithm, where each realization gets closer to the target density we are interested in. An essential property of any Markov chain is that the distribution of each draw only depends on the most recent value drawn before it. As we keep obtaining draws, each draw is probabilisticly more likely to have come from the target posterior distribution. Based on the convergence properties of MCMC, we can expect each draw to be coming from the real distribution of the parameter we are interested, once we have let the Markov chain run long enough.

- MCMC is the only suitable method when we don't have a conjugate prior.

### Metropolis-Hastings

@@

The beauty of the Gibbs sampling strategy is that, no matter how complex the space of the posterior distribution is, sampling from these high-dimensional spaces can be broken down into more straightforward draws from one-dimensional distributions. The downside of this approach is that, in high-dimensional settings, it may take a very long time to sample from each conditional one-dimensional distribution in order to reach convergence. Furthermore, because the Gibbs sampler uses a random-walk behavior, if several variables are highly correlated, traversing the space can take a very high amount of iterations because the step-size needs to be small in order to not step outside the target distribution.

@@

\textit{\textbf{Gamma}}: The $Gam(\alpha,\lambda)$ parameterization, where $\alpha$ is a shape parameter and $\lambda$ is a rate parameter, is used by \texttt{OpenBUGS}, \texttt{JAGS}, \texttt{Nimble}, and \texttt{greta}. The $Gam(\alpha,\beta)$ parameterization is used by \texttt{STAN}, using the scale parameter $\beta = 1 / \lambda$.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = FALSE)
library(knitr)
library(png)
```

@@

In Bayesian statistical inference, we are interested in finding the posterior distribution of certain parameters, given a prior distribution and a likelihood for those parameters our data. Much of Bayesian theory is built upon the definition of conditional probability, which states that we can find the probability of a variable $\theta$, conditioned on $x$, by $$ p(\theta | x) = \frac{p(x | \theta) p(\theta)}{p(x)}.$$ In any Bayesian inference, the first step is to find the joint probability distribution of all the observed and unobserved quantities in the problem. Using the definition of conditional probability, we can then find the posterior distribution of the parameters of interest. Bayes theorem states that the posterior distribution of a parameter $\theta$ is equal to $$p(\theta | x_1, x_2, \dots, x_n) = \frac{\ell(\theta | x_1, x_2, \dots, x_n) \pi(\theta)}{\int{\ell(\theta | x_1, x_2, \dots, x_n) \pi(\theta) d \theta}} = \frac{\ell(\theta | \textbf{x}) \pi(\theta)}{\int{\ell(\theta | \textbf{x}) \pi(\theta) d \theta}},$$ where $\ell(\theta | \textbf{x})$ is proportional to the likelihood of $\theta$, conditioned on the data, and $\pi(\theta)$ is the prior distribution of $\theta$.

For example, suppose that $X_i \sim \text{Bern}(\theta)$, and we are interested in inference on $\theta$. A reasonable prior to assign to $\theta$ would be a $\text{Beta}(\alpha, \beta)$, distribution, given that $\theta$ is constrained to be between 0 and 1, which is a property of the Beta distribution. Therefore, given that $\theta \sim \text{Beta}(\alpha, \beta)$ for some known values of $\alpha$ and $\beta$, then 
\begin{align*}
  p(\theta | x) &= \frac{\ell(\theta | \textbf{x}) \pi(\theta)}{\int{\ell(\theta | \textbf{x}) \pi(\theta) d \theta}} \\
  &= \frac{\theta ^ {\sum_{i=1}^{n} x_i} (1 - \theta) ^ {n - \sum_{i=1}^{n} x_i} \frac{\Gamma (\alpha + \beta)}{\Gamma(\alpha) + \Gamma(\beta)} \theta^{\alpha - 1} (1 - \theta)^{\beta - 1}}{\int \theta ^ {\sum_{i=1}^{n} x_i} (1 - \theta) ^ {n - \sum_{i=1}^{n} x_i} \frac{\Gamma (\alpha + \beta)}{\Gamma(\alpha) + \Gamma(\beta)} \theta^{\alpha - 1} (1 - \theta)^{\beta - 1} d \theta} \\
  &\propto \theta ^ {\sum (x_i) + \alpha - 1} (1 - \theta) ^ {n - \sum (x_i) + \beta - 1},
\end{align*}
which is the kernel of a $\text{Beta}\left(\sum (x_i) + \alpha, n - \sum (x_i) + \beta \right)$ distribution. Therefore, we can analytically calculate the posterior distribution of $\theta$ as $$p(\theta | x) \sim \text{Beta}\left(\sum (x_i) + \alpha, n - \sum (x_i) + \beta\right).$$

One challenge in bayesian inference comes in evaluating the integral in the denominator of Bayes theorem, which is used to normalize the posterior distribution. If we don't have distributions for the likelihood and prior that are conjugate (or in other words, play nice together), then it is impossible to analytically calculate the posterior distribution.

## Bayesian Computational Methods

The need for bayesian computational methods arises in our search for an alternative approach to finding the posterior if we cannot do so analytically. Most of the time, the goal of any bayesian analysis is to not to find the posterior distribution, but to calculate expectations with respect to the posterior. However, in order for these estimates to be accurate, we often need to know about the full behavior of that posterior distribution. In bayesian computing, the aim is to somehow obtain individual draws of values that would come from the posterior, and then aggregate a lot of them to compute the expectations we need.

@@

Markov Chain Monte Carlo (referred to as MCMC), is a common approach that allows for drawing samples from a target distribution, which in our case is the posterior. MCMC is an iterative algorithm, where we obtain one draw at a time from the parameter space from which we are trying to sample. In any Markov chain, the marginal distribution of each draw is conditioned on the values of our chain from the previous draw. As we keep obtaining draws, these marginal distributions asymptotically converge to the true target distribution. Therefore, if we run the markove chain for many iterations, we can expect to see draws coming from the true posterior distribution we are trying to estimate. MCMC is the only suitable method for obtaining the posterior distribution when we don't have conjugate priors.

### Metropolis-Hastings Algorithm

@@

The beauty of the Gibbs sampling strategy is that, no matter how complex the space of the posterior distribution is, sampling from these high-dimensional spaces can be broken down into more straightforward draws from one-dimensional distributions. Gibbs sampling is also easier to understand and implement, which makes it very versatile. However, full conditional distributions for each draw are needed in order to do the method.

Because the Gibbs sampler is a random-walk algorithm, if several variables are highly correlated, traversing the space can take a very high amount of iterations because the step-size needs to be small in order to not step outside the target distribution. A two-dimensional example of this is seen in Figure 1, where the joint distribution of $\theta_1$ and $\theta_2$ is highlgy correlated. Since each Markov transition only allows us to move in the direction of one dimension at a time, it may take many steps in order to travel up and down the parameter space, as demonstrated by the red arrows. In reality, the movement of the chains will be much more random since the algorithm does not know which direction will allow the chain to most efficiently travel through the posterior.

```{r, echo = FALSE, eval = TRUE, out.width = "50%"}
include_graphics("Paper-Figures/Gibbs1.png")
include_graphics("Paper-Figures/Gibbs2.png")
```

![\textit{This is a demonstration of a Gibbs sampler in a highly correlated two-dimensional parameter space. On the left, the ideal behavior of the markov chain is shown, moving as quickly as possible from the bottom-left to the top-right to efficiently sample from the target distribution without stepping outside of it. On the right, we see the more realistic behavior of the markov chain, as the random-walk behavior of the Gibbs sampler moves the chain around in directions that aren't always preferred.}](Paper-Figures/Blank.png)

@@

\textit{\textbf{Gamma}}: The $Gam(\alpha,\lambda)$ parameterization, where $\alpha$ is a shape parameter and $\lambda$ is a rate parameter, is used by all five implementations.
